---
title: "Data Visualization with ggplot2"
teaching: 60
exercises: 30
questions:
- ""
- ""
objectives:
- "Describe the role of data, aesthetics, and geoms in ggplot functions."
- "Choose the correct aesthetics and alter the geom parameters for a scatter plot, histogram, box plot, or line plot."
- "Apply additional aesthetics and geoms to a plot."
- "Customize plot scales, titles, subtitles, themes, fonts, layout, and orientation."
- "Apply a facet a plot."
- "Apply additional ggplot2-compatible plotting libraries."
- "Save a ggplot to a file."
- "List several resources for getting help with ggplot."
- "List several resources for creating informative scientific plots."

keypoints:
- ""
- ""
- ""
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("05-")
```


```{r, echo=FALSE}
submission_metadata <- read.csv("../data/submission_metatata_cleaned.csv")
```

We start by loading the required packages. **`ggplot2`** is included in the **`tidyverse`** package.

```{r load-package, message=FALSE, purl=FALSE}
library(tidyverse)
```

If not still in the workspace, load the data we saved in the previous lesson.


```{r load-data}
submission_metadata <- read.csv("../data/submission_metatata_cleaned.csv")
```

## Plotting with **`ggplot2`**

**`ggplot2`** is a plotting package that makes it simple to create complex plots
from data in a data frame. It provides a more programmatic interface for
specifying what variables to plot, how they are displayed, and general visual
properties. Therefore, we only need minimal changes if the underlying data change
or if we decide to change from a bar plot to a scatterplot. This helps in creating
publication quality plots with minimal amounts of adjustments and tweaking.

**`ggplot2`** functions like data in the 'long' format, i.e., a column for every dimension,
and a row for every observation. Well-structured data will save you lots of time
when making figures with **`ggplot2`**

ggplot graphics are built step by step by adding new elements. Adding layers in
this fashion allows for extensive flexibility and customization of plots.

To build a ggplot, we will use the following basic template that can be used for different types of plots:

ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()

- use the `ggplot()` function and bind the plot to a specific data frame using the
`data` argument

```{r, eval=FALSE, purl=FALSE}
ggplot(data =metadata)
```

- define a mapping (using the aesthetic (`aes`) function), by selecting the variables to be plotted and specifying how to present them in the graph, e.g. as x/y positions or characteristics such as size, shape, color, etc.

```{r, eval=FALSE, purl=FALSE}
ggplot(data =metadata, aes(x = sample, y = generation))
```

- add 'geoms' – graphical representations of the data in the plot (points,
lines, bars). **`ggplot2`** offers many different geoms; we will use some
common ones today, including:

* `geom_point()` for scatter plots, dot plots, etc.
* `geom_boxplot()` for, well, boxplots!
* `geom_line()` for trend lines, time series, etc.

To add a geom to the plot use the `+` operator. Because we have two continuous variables,
let's use `geom_point()` first:

```{r first-ggplot, purl=FALSE}
ggplot(data =metadata, aes(x = sample, y = generation))+
geom_point()
```

The `+` in the **`ggplot2`** package is particularly useful because it allows you
to modify existing `ggplot` objects. This means you can easily set up plot
templates and conveniently explore different types of plots, so the above
plot can also be generated with code like this:

```{r, first-ggplot-with-plus, eval=FALSE, purl=FALSE}
# Assign plot to a variable
metadata_plot <- ggplot(data =metadata, aes(x = sample, y = generation))

# Draw the plot
metadata_plot +
geom_point()
```

```{r, eval=FALSE, purl=TRUE, echo=FALSE, purl=FALSE}
# Create a ggplot and draw it
metadata_plot <- ggplot(data =metadata, aes(x = sample, y = generation))
metadata_plot  +
geom_point()
```

**Notes**

- Anything you put in the `ggplot()` function can be seen by any geom layers
that you add (i.e., these are universal plot settings). This includes the x- and
y-axis mapping you set up in `aes()`.
- You can also specify mappings for a given geom independently of the
mappings defined globally in the `ggplot()` function.
- The `+` sign used to add new layers must be placed at the end of the line containing
the *previous* layer. If, instead, the `+` sign is added at the beginning of the line
containing the new layer, **`ggplot2`** will not add the new layer and will return an
error message.

```{r, ggplot-with-plus-position, eval=FALSE, purl=FALSE}
# This is the correct syntax for adding layers
metadata_plot  +
geom_point()

# This will not add the new layer and will return an error message
metadata_plot
+ geom_point()
```

> ## Challenge (optional)
>
> Scatter plots can be useful exploratory tools for small datasets. For data
> sets with large numbers of observations, such as the `metadata` data
> set, overplotting of points can be a limitation of scatter plots. One strategy
> for handling such settings is to use hexagonal binning of observations. The
> plot space is tessellated into hexagons. Each hexagon is assigned a color
> based on the number of observations that fall within its boundaries.  To use
> hexagonal binning with **`ggplot2`**, first install the R package `hexbin`
> from CRAN:
>
>
> ```{r, eval = FALSE}
> install.packages("hexbin")
> library(hexbin)
> ```
>
> Then use the `geom_hex()` function:
>
> ```{r, eval = FALSE}
> surveys_plot +
>  geom_hex()
> ```
>
> - What are the relative strengths and weaknesses of a hexagonal bin plot
>   compared to a scatter plot? Examine the above scatter plot and compare it
>   with the hexagonal bin plot that you created.
{: .challenge}

## Building your plots iteratively

Building plots with **`ggplot2`** is typically an iterative process. We start by
defining the dataset we'll use, lay out the axes, and choose a geom:

```{r create-ggplot-object, purl=FALSE}
ggplot(data =metadata, aes(x = sample, y = generation)) +
geom_point()
```

Then, we start modifying this plot to extract more information from it. For
instance, we can add transparency (`alpha`) to avoid overplotting:

```{r adding-transparency, purl=FALSE}
ggplot(data =metadata, aes(x = sample, y = generation)) +
geom_point(alpha = 0.1)
```

We can also add colors for all the points:

```{r adding-colors, purl=FALSE}
ggplot(data =metadata, aes(x = sample, y = generation)) +
geom_point(alpha = 0.1, color = "blue")
```

Or to color each strain in the plot differently, you could use a vector as an input to the argument **color**. **`ggplot2`** will provide a different color corresponding to different values in the vector. Here is an example where we color with **`strain`**:


```{r color-by-strain, purl=FALSE}
ggplot(data =metadata, aes(x = sample, y = generation)) +
geom_point(alpha = 0.1, aes(color = strain))
```

> ## Challenge
>
> Use what you just learned to create a scatter plot of `genome_size` over
> `strain` with the plot types showing in different colors. Is this a good
> way to show this type of data?
>
>> ## Solution
>> ```{r scatter-challenge, answer=TRUE, purl=FALSE}
>> ggplot(data = metadata, aes(x = strain, y = genome_size)) +
>>    geom_point(aes(color = strain))
>> ```
> {: .solution}
{: .challenge}

## Boxplot

We can use boxplots to visualize the distribution of genome_size within each strain:

```{r boxplot, purl=FALSE}
ggplot(data =metadata, aes(x = sample, y = generation))+
geom_boxplot()
```

By adding points to boxplot, we can have a better idea of the number of
measurements and of their distribution:

```{r boxplot-with-points, purl=FALSE}
ggplot(data =metadata, aes(x = sample, y = generation))+
geom_boxplot(alpha = 0) +
geom_jitter(alpha = 0.3, color = "tomato")
```

Notice how the boxplot layer is behind the jitter layer? What do you need to
change in the code to put the boxplot in front of the points such that it's not
hidden?

> ## Challenge
>
> Boxplots are useful summaries, but hide the *shape* of the distribution. For
> example, if the distribution is bimodal, we would not see it in a
> boxplot. An alternative to the boxplot is the violin plot, where the shape
(of the density of points) is drawn.
>
> a) Replace the box plot with a violin plot; see `geom_violin()`.
>
> In many types of data, it is important to consider the *scale* of the
> observations.  For example, it may be worth changing the scale of the axis to
> better distribute the observations in the space of the plot.  Changing the scale
> of the axes is done similarly to adding/modifying other components (i.e., by
> incrementally adding commands). Try making these modifications:
>
> b) Represent genome_size on the log~10~ scale; see `scale_y_log10()`.
>
> So far, we've looked at the distribution of genome_size within strain. Try making a
> new plot to explore the distribution of another variable within each strain.
>
> c) Create a boxplot for `hindfoot_length`. Overlay the boxplot layer on a jitter
>   layer to show actual measurements.
>
> d) Add color to the data points on your boxplot according to the plot from which
>   the sample was taken (`plot_id`).
>
> *Hint:* Check the class for `plot_id`. Consider changing the class of `plot_id` from integer to factor. Why does this change how R makes the graph?
{: .challenge}


## Plotting time series data

Let's calculate number of strains per each generation. First we need
to group the data and strain records within each group:

```{r, purl=FALSE}
sample_per_generation <- metadata%>%
group_by(strain,generation) %>%
tally()
```

Time series data can be visualized as a line plot with strains on the x axis and generation
on the y axis:

```{r first-time-series, purl=FALSE}
ggplot(data =sample_per_generation, aes(x =generation, y = n)) +
geom_line()
```


```{r final-challenge, eval=FALSE, purl=TRUE, echo=FALSE}
### Final plotting challenge:
##  With all of this information in hand, please take another five
##  minutes to either improve one of the plots generated in this
##  exercise or create a beautiful graph of your own. Use the RStudio
##  ggplot2 cheat sheet for inspiration:
##  https://www.rstudio.com/wp-content/uploads/2015/08/ggplot2-cheatsheet.pdf
```