---
title: "Data Visualization with ggplot2"
teaching: 60
exercises: 30
questions:
- "How do I plot my data in R?"
- "How do I adjust plots to better display my data?"
objectives:
- "Describe the role of data, aesthetics, and geoms in ggplot functions."
- "Choose aesthetics and alter the geom parameters for a scatter plot, histogram, or box plot."
- "Customize plot scales, titles, subtitles, themes, fonts, layout, and orientation."
- "Apply a facet to a plot."
- "Save a ggplot to a file."
- "List resources for getting help with ggplot and creating informative scientific plots."

keypoints:
- "ggplot2 maps variables in a tidy data frame to visual aesthetics such as position and color."
- "Plots are built iteratively from different components: data, geoms, scales, and themes."
- "You can customize any aspect of a plot to best display your data."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("05-")
```

```{r, echo = FALSE, eval = TRUE, purl = FALSE}
## silently read in CSV file from FigShare

variants <- read.csv("https://ndownloader.figshare.com/files/14632895")
```

We start by loading the package **`ggplot2`**. 

```{r load-ggplot2, message=FALSE, purl = TRUE}
library(ggplot2)
```

The *gg* in **`ggplot2`** stands for "grammar of graphics". The idea is that 
any tidy data set can be made into an informative plot in a systematic way, by mapping
variables in the data to positions, colours, shapes, and sizes of visual output. 
These visual features are referred to as "aesthetics", or "aes" for short.
There are different ways of mapping aesthetics to visual features - 
bar plots, line plots, and so on. These differ in their geometry and so are referred 
to as "geometries", or "geom" for short.

Some simple examples will help. We will first plot one of R's built-in datasets,
the [*Iris* flower dataset](https://en.wikipedia.org/wiki/Iris_flower_data_set),
which measured sizes of iris flowers in different species, in 1930s Canada^[The iris dataset was collected by scientists who supported eugenics, a discredited and discriminatory historical practice. More background is given on the wikipedia page.].
This data set records the length and width of two different parts of flowers, the Sepal 
and the Petal, for three different species, (*Iris setosa*, *Iris virginica* and *Iris versicolor*).

```{r print-iris, message=FALSE, purl = TRUE}
head(iris)
```

This scatter plot displays points (`geom_point`), where:

* x-axis position is given by `Petal.Width`
* y-axis position is given by `Petal.Length`
* colour is given by `Species`

```{r scatter-plot-iris, message=FALSE, purl = TRUE, fig.height=3, fig.width=4.5}
ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length, colour = Species)) +
  geom_point()
```

We could just as easily make a scatter plot where:

* x-axis position is given by `Species`
* y-axis position is given by `Petal.Length`
* colour is given by `Petal.Width`

```{r scatter-plot-iris-ugly, message=FALSE, purl = TRUE, fig.height=3, fig.width=4.5}
ggplot(data = iris, aes(x = Species, y = Petal.Length, colour = Petal.Width)) +
  geom_point()
```

You may find this plot less informative, or even uglier. This makes the point that
the choice of how to map data features onto visual aesthetics is important.

## Plotting with **`ggplot2`**

**`ggplot2`** is a plotting package that makes it simple to create complex plots
from a tidy data frame. It provides a programmatic interface for
specifying what variables to plot, how they are displayed, and other visual
properties. This helps in creating
publication quality plots with minimal amounts of adjustments and tweaking.

**`ggplot2`** works with tidy data, i.e., a column for every variable,
and a row for every observation. Well-structured data will save you lots of time
when making figures with **`ggplot2`**. If you need to reorganise your
data, you can use other tidyverse packages such as **`dplyr`** and **`tidyr`**, 
and then afterwards plot the reorganised data using **`ggplot2`**.

ggplot graphics are built step by step by adding new elements. Adding layers in
this fashion allows for extensive flexibility and customization of plots.

To build a ggplot, we will use the following basic template that can be used for different types of plots:

```{r, eval=FALSE}
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()
```

The `ggplot()` function sets up an empty plot, which is "bound" to a specific data frame, using the
      `data` argument. We will use the `variants` data frame from earlier in this workshop.

```{r, eval=FALSE, purl = TRUE}
ggplot(data = variants)
```

The aesthetic (`aes`) function defines a mapping by selecting the variables to be
plotted and specifying how to present them in the plot. 
Here we use genomic position `POS` for the `x` position, and read depth `DP` for
the `y` position.

```{r, eval=FALSE, purl = TRUE}
ggplot(data = variants, aes(x = POS, y = DP))
```

'geoms' provide graphical representations of the data in the plot. 
**`ggplot2`** offers many different geoms; we will use some
  common ones today, including:

      * `geom_point()` for scatter plots, dot plots, etc.
      * `geom_boxplot()` for, well, boxplots!
      * `geom_line()` for trend lines, time series, etc.

To add a geom to the plot we use the `+` operator. 
Because we have two continuous variables, let's use `geom_point()` first:

```{r first-ggplot, purl = TRUE}
ggplot(data = variants, aes(x = POS, y = DP)) +
  geom_point()
```

The `+` in the **`ggplot2`** package allows you to modify existing `ggplot` objects,
by adding more plot elements. This means you can easily set up plot
templates and conveniently explore different types of plots, like this:

```{r, first-ggplot-with-plus, eval=FALSE, purl = TRUE}
# Assign plot to a variable
coverage_plot <- ggplot(data = variants, aes(x = POS, y = DP))

# Draw the plot
coverage_plot +
    geom_point()
```

**Notes**

- Data and aesthetics in the `ggplot()` function can be seen by geom layers
  that you add (i.e., these are universal plot settings). This includes the x- and
  y-axis mapping you set up in `aes()`.
- You can instead specify mappings for a given geom to replace or add to those
  mappings defined globally in the `ggplot()` function.
- The `+` sign used to add new layers must be placed at the end of the line containing
the *previous* layer. This is because R assumes that the end of a line is the end of
a command, unless you tell it otherwise. So, if the `+` sign is added at the beginning of 
the next line containing the new layer, **`ggplot2`** will not add that new layer and will 
return an error message.

```{r, ggplot-with-plus-position, eval=FALSE, purl = TRUE}
# This is the correct syntax for adding layers
coverage_plot +
  geom_point()

# This will not add the new layer and will return an error message
coverage_plot
  + geom_point()
```

## Building your plots iteratively: geoms.

Building plots with **`ggplot2`** is typically an iterative process. We start by
defining the dataset we'll use, lay out the axes, and choose a geom:

```{r create-ggplot-object, purl = TRUE}
ggplot(data = variants, aes(x = POS, y = DP)) +
  geom_point()
```

To be clear, this plot tells us the position and read depth of each genomic variant.

To learn more, we start modifying this plot. For example nearby points may be 
"overplotted" so that it is hard to distinguish one point from many nearby points.
To reduce overplotting, we can add transparency using the aesthetic (`alpha`):

```{r adding-transparency, purl = TRUE}
ggplot(data = variants, aes(x = POS, y = DP)) +
    geom_point(alpha = 0.5)
```

## Using colors as an aesthetic

If plotting with a different color is more clear, we can add colors for all the points:

```{r adding-colors, purl = TRUE}

ggplot(data = variants, aes(x = POS, y = DP)) +
  geom_point(alpha = 0.5, color = "blue")
```

To add sample-specific visual information, we can color each sample in the plot differently.
**`ggplot2`** is designed to make it easy to map variables to visual aesthetics, such as **color**. 
Here is an example where we color with **`sample_id`**:

```{r color-by-sample-1, purl = TRUE}
ggplot(data = variants, aes(x = POS, y = DP, color = sample_id)) +
  geom_point(alpha = 0.5)
```

> ## Challenge: 
>
> 1. What is the difference between `color = sample_id` (no quotes) and `color = "sample_id"` (auotes)? 
> 2. What is the difference between specifying color **inside** the `aes()` function, or **outside**?
>
>> ## Solution
>> 
>> ```{r, echo=TRUE}
>> ggplot(data = variants, aes(x = POS, y = DP, color = "sample_id")) +
>>  geom_point(alpha = 0.5)
>> ```
>> 
>> In this case, `ggplot` interprets `color = "sample_id"` as mapping the *string* `"sample_id"`,
>> not the variable `sample_id`. These distinctions turn out to be important.
>>
> {: .solution}
{: .challenge}

We can plot the same aesthetic mappings using different geometries. For example, 
we can change the geom from point to column, and positions and colors will be still 
determined in the same way, but they will look different:

```{r color-by-sample-2, purl = TRUE}
ggplot(data = variants, aes(x = POS, y = DP, color = sample_id)) +
  geom_col(alpha = 0.5)
```

To find out more about the different aesthetics, you can use the help functions, such as 
`?geom_point` and `?geom_col`. There are many more useful resources that we link to below.


## Building plots iteratively: axis labels

To make our plot more interpretable, we can add axis labels:

```{r add-axis-labels, purl = TRUE}
ggplot(data = variants, aes(x = POS, y = DP, color = sample_id)) +
  geom_jitter(alpha = 0.5) +
  labs(x = "Base Pair Position",
       y = "Read Depth (DP)")
```

ggplot also allows you to change the way in which variables are mapped to
aesthetics, using **scales**. For example, variables that are widely dispersed 
may be easier to visualise on a log scale, where equal spacings on the page 
represent equal fold-changes in the variable. Here we use the built-in
`scale_y_log10` to display the y-axis on a log(10) scale.

```{r y-log10-scale, purl = TRUE}
ggplot(data = variants, aes(x = POS, y = DP, color = sample_id)) +
  geom_jitter(alpha = 0.5) +
  labs(x = "Base Pair Position",
       y = "Read Depth (DP)") +
  scale_y_log10()
```

Although we could transform the x-axis using `scale_x_log10`, that is not appropriate if the x-axis represents genomic position.
ggplot allows you to customize almost any aspect of a plot, whether or not it is a good idea!

To find out more about different scales, you can again use the help functions, such as 
`?scale_y_log10`. Or, type `scale_` and tab autocomplete to find out what other options are available.

> ## Challenge
>
> Use what you just learned to create a scatter plot of mapping quality (`MQ`) over
> position (`POS`) with the samples showing in different colors. Make sure to give your plot
> relevant axis labels.
> 
> Does it make sense to try a different y-axis scale here?
>
> > ## Solution
> > ```{r scatter-challenge, purl = TRUE}
> >  ggplot(data = variants, aes(x = POS, y = MQ, color = sample_id)) +
> >   geom_point() +
> >   labs(x = "Base Pair Position",
> >        y = "Mapping Quality (MQ)")
> > ```
> {: .solution}
{: .challenge}

## Creating subplots by Faceting

**`ggplot2`** has a feature called *faceting* that allows the user to split one
plot into multiple plots based on a factor included in the dataset. This offers
an alternative to splittting differnt factors by, for example, colour or shape.
We will use it to split our read depth plot into three panels, one for each sample.

```{r first-facet, purl = TRUE}
ggplot(data = variants, aes(x = POS, y = DP, color = sample_id)) +
 geom_point() +
 labs(x = "Base Pair Position",
      y = "Read Depth (DP)") +
 facet_grid(. ~ sample_id)
```

This looks ok, but it would be easier to read if the plot facets were stacked vertically rather
than horizontally. The `facet_grid`
geometry allows you to explicitly specify how you want your plots to be
arranged via formula notation (`rows ~ columns`; a `.` can be used as
a placeholder that indicates only one row or column).

```{r second-facet, purl = TRUE}
ggplot(data = variants, aes(x = POS, y = DP, color = sample_id)) +
 geom_point() +
 labs(x = "Base Pair Position",
      y = "Read Depth (DP)") +
 facet_grid(sample_id ~ .)
```

In this case, stacking the plots vertically makes sense as all samples
measure read depth along the same genome, represented by the x-axis.

## Building plots iteratively: adjusting plot appearance with themes

Plots have many visual elements besides the data - axis and tick labels, grid lines, 
background colours, and so on. All of these can vary in their appearance. Text elements
vary in font. Most elements vary by their size and color, and even by their presence or
absence. ggplot controls the appearance of these other plot elements with **themes**. 

For example, plots with white background look more readable when printed.  We can set
the background to white using built-in theme, `theme_bw()`:

```{r facet-plot-white-bg, purl = TRUE}
ggplot(data = variants, aes(x = POS, y = MQ, color = sample_id)) +
  geom_point() +
  labs(x = "Base Pair Position",
       y = "Mapping Quality (MQ)") +
  facet_grid(sample_id ~ .) +
  theme_bw()
```

We can remove individual plot elements that aren't needed, for example the grid lines:

```{r facet-plot-white-bg-no-grid, purl = TRUE}
ggplot(data = variants, aes(x = POS, y = MQ, color = sample_id)) +
  geom_point() +
  labs(x = "Base Pair Position",
       y = "Mapping Quality (MQ)") +
  facet_grid(sample_id ~ .) +
  theme_bw() +
  theme(panel.grid = element_blank())
```

> ## Challenge
>
> Use what you just learned to create a scatter plot of PHRED scaled quality (`QUAL`) over
> position (`POS`) with the samples showing in different colors. Make sure to give your plot
> relevant axis labels.
>
> > ## Solution
> > ```{r scatter-challenge-2, purl = TRUE}
> >  ggplot(data = variants, aes(x = POS, y = QUAL, color = sample_id)) +
> >   geom_point() +
> >   labs(x = "Base Pair Position",
> >        y = "PHRED-sacled Quality (QUAL)") +
> >   facet_grid(sample_id ~ .)
> > ```
> {: .solution}
{: .challenge}

In addition to `theme_bw()`, which changes the plot background to white, **`ggplot2`**
comes with several other themes which can be useful to quickly change the look
of your visualization. The complete list of built-in themes is available
at <https://ggplot2.tidyverse.org/reference/ggtheme.html>. `theme_minimal()` and
`theme_light()` are popular, and `theme_void()` can be useful as a starting
point to create a new hand-crafted theme.

The
[ggthemes](https://jrnold.github.io/ggthemes/reference/index.html) package
provides a wide variety of options (including an Excel 2003 theme).
The [**`ggplot2`** extensions website](https://exts.ggplot2.tidyverse.org/) provides a list
of packages that extend the capabilities of **`ggplot2`**, including additional
themes.

## Making barplots to count the number of observations in groups

Plots with similar appearances can mean different things.
Earlier we used `geom_col` to plot read depth (DP) as the height of a column.
However, bar plots are commonly used to count the number of observations in groups.
ggplot distinguishes between `geom_col`, which plots the raw dat supplied, and `geom_bar`, 
which does the counting for you. 

We can create barplots of counts using `geom_bar`: Let's make a barplot showing
the number of variants for each sample that are indels.

```{r barplot, purl = TRUE}
ggplot(data = variants, aes(x = INDEL)) +
  geom_bar() +
  facet_grid(sample_id ~ .)
```

> ## Challenge
> Since we already have the sample_id labels on the individual plot facets, we don't need the
> legend. Use the help file for `geom_bar` and any other online resources you want to use to
> remove the legend from the plot.
>
>> ## Solution
>> ```{r barplot-challenge, purl = TRUE}
>> ggplot(data = variants, aes(x = INDEL, color = sample_id)) +
>>    geom_bar(show.legend = F) +
>>    facet_grid(sample_id ~ .)
>> ```
> {: .solution}
{: .challenge}

There is much more flexibility in terms of statistical summaries and counting in ggplot.
See the Tip box "Where to learn more about plotting" to find out more.


## Exporting plots to files

To export a plot to a file, use the **Export** button, in the bottom right plotting window in RStudio.
Or you can write this into a script using the function `ggsave` - remember, use `?ggsave` to
ask for help from your R session.

> ## Challenge
>
> With all of this information in hand, please take another five minutes to
> either improve one of the plots generated in this exercise or create a
> beautiful graph of your own. Use the RStudio [**`ggplot2`** cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf)
> for inspiration. Here are some ideas:
>
> * See if you can change the size or shape of the plotting symbol.
> * Can you find a way to change the name of the legend? What about its labels?
> * Try using a different color palette (see
>   http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/).
{: .challenge}

> ## Tip: Where to learn more about plotting
> There are amazing online and print resources for learning more about ggplot2.
> Some of them talk about the ideas behind plotting, and give galleries of examples.
> Others are more practical about how to program with ggplot2.
> - [R for ecology ggplot2](https://datacarpentry.org/R-ecology-lesson/04-visualization-ggplot2.html):
    Episode from Data Carpentry lesson introducing ggplot with a different example dataset.
> - [ggplot2: elegant graphics for data analysis](https://ggplot2-book.org/): By Hadley Wickham, 
    Danielle Navarro, and Thomas Lin Pedersen. Gives some basics of ggplot2, but its primary focus
    is explaining the Grammar of Graphics that ggplot2 uses.
> - [Data Visualization: A Practical Introduction](https://kieranhealy.org/publications/dataviz/): 
    By Kieran Healy. Teaches how to makedata visualizations with R and ggplot2 in a clear, 
    sensible, and reproducible way. Accompanied by the `socviz` package with example code.
> - [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/): By Claus Wilke.
    A guide to making visualizations that accurately reflect the data, tell a story, and look 
    professional. Complete source code, with plots made in ggplot2, is available.
{: .callout}
