---
title: "Data Visualization with ggplot2"
teaching: 60
exercises: 30
questions:
- ""
- ""
objectives:
- ""
- "Describe the role of data, aesthetics, and geoms in ggplot functions."
- "Choose the correct aesthetics and alter the geom parameters for a scatter plot, histogram, or box plot."
- "Layer multiple geometries in a single plot."
- "Customize plot scales, titles, subtitles, themes, fonts, layout, and orientation."
- "Apply a facet to a plot."
- "Apply additional ggplot2-compatible plotting libraries."
- "Save a ggplot to a file."
- "List several resources for getting help with ggplot."
- "List several resources for creating informative scientific plots."

keypoints:
- ""
- ""
- ""
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("05-")
```

```{r, echo = FALSE, eval = TRUE, purl = FALSE}
## silently read in CSV file from FigShare

variants <- read.csv("https://ndownloader.figshare.com/files/14632895")
```

We start by loading the required packages. **`ggplot2`** is included in the **`tidyverse`** package.

```{r load-package, message=FALSE, purl=FALSE}
library(tidyverse)
```

## Plotting with **`ggplot2`**

**`ggplot2`** is a plotting package that makes it simple to create complex plots
from data in a data frame. It provides a more programmatic interface for
specifying what variables to plot, how they are displayed, and general visual
properties. Therefore, we only need minimal changes if the underlying data change
or if we decide to change from a bar plot to a scatterplot. This helps in creating
publication quality plots with minimal amounts of adjustments and tweaking.

**`ggplot2`** functions like data in the 'long' format, i.e., a column for every dimension,
and a row for every observation. Well-structured data will save you lots of time
when making figures with **`ggplot2`**

ggplot graphics are built step by step by adding new elements. Adding layers in
this fashion allows for extensive flexibility and customization of plots.

To build a ggplot, we will use the following basic template that can be used for different types of plots:

ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()

- use the `ggplot()` function and bind the plot to a specific data frame using the
`data` argument

```{r, eval=FALSE, purl=FALSE}
ggplot(data = variants)
```

- define a mapping (using the aesthetic (`aes`) function), by selecting the variables to be plotted and specifying how to present them in the graph, e.g. as x/y positions or characteristics such as size, shape, color, etc.

```{r, eval=FALSE, purl=FALSE}
ggplot(data = variants, aes(x = sample_id, y = generation))
```

- add 'geoms' – graphical representations of the data in the plot (points,
lines, bars). **`ggplot2`** offers many different geoms; we will use some
common ones today, including:

* `geom_point()` for scatter plots, dot plots, etc.
* `geom_boxplot()` for, well, boxplots!
* `geom_line()` for trend lines, time series, etc.

To add a geom to the plot use the `+` operator. Let's start out by
plotting the positions with observed variations for each sample using
`geom_point()`:

```{r first-ggplot, purl=FALSE}
ggplot(data = variants, aes(x = POS, y = sample_id))+
geom_point()
```

At a glance, we can see that the SRR2584866 sample has many more
positions of variation than the other two samples, but a lot of the points overlap with each other. To get a clearer picture of all
of the data, we can use the `geom_jitter()` geom.

```{r jitter-plot, purl=FALSE}
ggplot(data = variants, aes(x = POS, y = sample_id))+
geom_jitter()
```

The `+` in the **`ggplot2`** package is particularly useful because it allows you
to modify existing `ggplot` objects. This means you can easily set up plot
templates and conveniently explore different types of plots, so the above
plot can also be generated with code like this:

```{r, first-ggplot-with-plus, eval=FALSE, purl=FALSE}
# Assign plot to a variable
metadata_plot <- ggplot(data =metadata, aes(x = sample, y = generation))

# Draw the plot
metadata_plot +
geom_point()
```

**Notes**

- Anything you put in the `ggplot()` function can be seen by any geom layers
that you add (i.e., these are universal plot settings). This includes the x- and
y-axis mapping you set up in `aes()`.
- You can also specify mappings for a given geom independently of the
mappings defined globally in the `ggplot()` function.
- The `+` sign used to add new layers must be placed at the end of the line containing
the *previous* layer. If, instead, the `+` sign is added at the beginning of the line
containing the new layer, **`ggplot2`** will not add the new layer and will return an
error message.

```{r, ggplot-with-plus-position, eval=FALSE, purl=FALSE}
# This is the correct syntax for adding layers
variants_plot  +
geom_point()

# This will not add the new layer and will return an error message
variants_plot
+ geom_point()
```

## Building your plots iteratively

Building plots with **`ggplot2`** is typically an iterative process. We start by
defining the dataset we'll use, lay out the axes, and choose a geom:

```{r create-ggplot-object, purl=FALSE}
ggplot(data = metadata, aes(x = sample, y = generation)) +
geom_point()
```

Alternatively, we can make our data points slightly transparent. We
do this by adding an aesthetic within our `geom_point()` function 
call. The name of the aesthetic we want is "alpha". 

```{r alpha, purl=FALSE}
ggplot(data = variants, aes(x = POS, y = sample_id))+
geom_point(alpha = 0.1)
```

We can also add colors for all the points:

```{r adding-colors, purl=FALSE}
ggplot(data = variants, aes(x = POS, y = sample_id))+
geom_point(alpha = 0.1, color = "blue")
```

We could also color code each of our points according to the number of
reads (after filtering) that support that variant. This information
is in the `DP` column: 


```{r color-by-DP, purl=FALSE}
ggplot(data = variants, aes(x = POS, y = sample_id))+
geom_point(aes(color = DP))
```

> ## Challenge
>
> Use what you just learned to create a scatter plot of `DP` 
> versus
> `sample_id`. Is this a good
> way to show this type of data?
>
>> ## Solution
>> ```{r scatter-challenge, answer=TRUE, purl=FALSE}
>> ggplot(data = variants, aes(x = DP, y = sample_id)) +
>>    geom_point()
>> ```
> {: .solution}
{: .challenge}

## Boxplot

When we have one continuous and one categorical variable
it is usually much more appropriate to make a boxplot
than a scatterplot. In this case, our continuous variable
is read depth (`DP`) and our categorical variable in
`sample_id`. We can use boxplots to visualize the distribution of (filtered) sequencing depth `DP` within each sample:

```{r boxplot, purl=FALSE}
ggplot(data = variants, aes(x = sample_id, y = DP))+
geom_boxplot()
```

By adding points to boxplot, we can have a better idea of the number of
measurements and of their distribution:

```{r boxplot-with-points, purl=FALSE}
ggplot(data = variants, aes(x = sample_id, y = DP))+
geom_boxplot() +
geom_jitter(alpha = 0.3, color = "tomato")
```

> ## Exercise
> 
> In the plot above, the boxplot layer appears behind the 
> jitter layer. Recreate the plot, but put the boxplot 
> in front of the jitter plot so that it's easier to see
> the boxplot.
>> ## Solution
>> ```{r}
>> ggplot(data = variants, aes(x = sample_id, y = DP)) +
>> geom_jitter(alpha = 0.3, color = "tomato") + 
>> geom_boxplot()
>> ```
> {: .solution}
{: .challenge}

> ## Exercise (intermediate)
>
> Boxplots are useful summaries, but hide the *shape* of the distribution. For
> example, if the distribution is bimodal, we would not see it in a
> boxplot. An alternative to the boxplot is the violin plot, where the shape
(of the density of points) is drawn.
>
> Replace the box plot with a violin plot. You'll need to 
> look at the documenation to find an appropriate geom.
> 
>> ## Solution
>> ```{r}
>> ggplot(data = variants, aes(x = sample_id, y = DP)) +
>> geom_jitter(alpha = 0.3, color = "tomato") + 
>> geom_violin()
>> ```
> {: .solution}
{: .challenge}

> ## Exercise (intermediate)
> So far, we've looked at the distribution of read
> depth within each sample. Make a new plot to
> explore the distribution of another 
> continuous variable within each sample.
> 
>> ## Solution
>> One possible answer:
>> 
>> ```{r}
>> ggplot(data = variants, aes(x = sample_id, y = QUAL)) +
>> geom_jitter(alpha = 0.3, color = "tomato") + 
>> geom_boxplot()
>> ```
> {: .solution}
{: .challenge}

> c) Create a boxplot for `hindfoot_length`. Overlay the boxplot layer on a jitter
>   layer to show actual measurements.
>
> d) Add color to the data points on your boxplot according to the plot from which
>   the sample was taken (`plot_id`).
>
> *Hint:* Check the class for `plot_id`. Consider changing the class of `plot_id` from integer to factor. Why does this change how R makes the graph?
{: .challenge}
