---
title: "Exploring Data Frames"
teaching: 20
exercises: 10
questions:
- "How can I manipulate a data frame?"
objectives:
- "Be able to add and remove rows and columns."
- "Be able to remove rows with `NA` values."
- "Be able to append two data frames"
- "Be able to articulate what a `factor` is and how to convert between `factor` and `character`."
- "Be able to find basic properties of a data frames including size, class or type of the columns, names, and first few rows."
keypoints:
- "Use `cbind()` to add a new column to a data frame."
- "Use `rbind()` to add a new row to a data frame."
- "Remove rows from a data frame."
- "Use `na.omit()` to remove rows from a data frame with `NA` values."
- "Use `levels()` and `as.character()` to explore and manipulate factors"
- "Use `str()`, `nrow()`, `ncol()`, `dim()`, `colnames()`, `rownames()`, `head()` and `typeof()` to understand structure of the data frame"
- "Read in a csv file using `read.csv()`"
- "Understand `length()` of a data frame"
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("05-")
```


At this point, you've see it all - in the last lesson, we toured all the basic
data types and data structures in R. Everything you do will be a manipulation of
those tools. But a whole lot of the time, the star of the show is going to be
the data frame - the table that we created by loading information from a csv file. In this lesson, we'll learn a few more things
about working with data frames.

## Adding columns and rows in data frame

We learned last time that the columns in a data frame were vectors, so that our
data are consistent in type throughout the column. As such, if we want to add a
new column, we need to start by making a new vector:

```{r, echo = FALSE}
ecoli_meta <- read.csv("~/wkspace/r-novice-gapminder/data/Ecoli_metadata.csv")
```

```{r}
samp_n <- rep(1:6,5)
ecoli_meta <- cbind(ecoli_meta, samp_n)

```

We can then also add this as a column via:

```{r, error=TRUE}
ecoli_meta$samp_n2 <- samp_n
```


Now how about adding rows - in this case, we saw last time that the rows of a
data frame are made of lists:

```{r}
newRow <- ecoli_meta[30,]
ecoli_meta <- rbind(ecoli_meta, newRow)
```


## Removing rows
We can also remove rows, so let's drop the offending row.

```{r}
ecoli_meta[-31,]
```


Notice the comma with nothing after it to indicate we want to drop the entire fourth row.

Note: We could also remove both new rows at once by putting the row numbers
inside of a vector: `ecoli_meta[c(-4,-5),]`

Alternatively, we can drop all rows with `NA` values:

```{r}
ecoli_meta <- read.csv("~/wkspace/r-novice-gapminder/data/Ecoli_metadata.csv",na.strings = "N/A")

```

Let's reassign the output to `ecoli_meta`, so that our changes will be permanent:

```{r}
ecoli_meta <- na.omit(ecoli_meta)
```

## Appending to a data frame

The key to remember when adding data to a data frame is that *columns are
vectors or factors, and rows are lists.* We can also glue two data frames
together with `rbind`:

```{r}
ecoli_meta <- rbind(ecoli_meta,ecoli_meta)
ecoli_meta
```
There's a variety of functions that can give us some metadata about our dataframes
```{r}
rownames(ecoli_meta)
colnames(ecoli_meta)
dim(ecoli_meta)
ncol(ecoli_meta)
nrow(ecoli_meta)
```

Lastly we can use these to reassign the values.
```{r}
colnames(ecoli_meta) <- c("X1","X2","X3","X4","X5","X6","X7")
```


> ## Challenge 1
>
> Rename the columns with your own column names
